Random Walks across Social Networks: The Markovian Movement of a Stochastic Soccerball

### 2020-07-25 rmd/Preprocessing_Python.Rmd
done

---
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.4.2
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
---

@@

```{python}
for league in national_leagues:
    preprocess(league)
```
``

>>>>

---
jupyter:
  jupytext:
    formats: nb//ipynb,rmd//Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.4.2
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
---

@@

```{python}
if __name__ == '__main__':
    for league in national_leagues:
        preprocess(league)
```

### 2020-07-25 rmd/SocialNetworkAnalysis_R.Rmd
done

---
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.4.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

@@

```{r}
TabulateSequence <- function(events, possession_sequence) {
    # Return a markdown table starting from the row before a possession sequence begins, and continuing until it ends.

    start <- min(events[events$possession == possession_sequence,]$X)
    stop <- max(events[events$possession == possession_sequence,]$X) + 1
    table <- events[start:stop,]
    df <- data.frame(table$matchPeriod, table$time, table$team, table$source, table$subEventName)
    names(df) <- c('Half', 'Time', 'Team', 'Event Description', 'Player')

@@

    d <- read.csv('../tables/MXG.csv')
    top <- d[d$Degree>1000,]
    rownames(top) <- c()
    top['Loops/Degree'] <- top$Loops/top$Degree
    top['Goals/Degree'] <- top$Goals/(top$Degree)
    t <- top[c('Player','Team','Degree','Goals',
               'Goals/Degree','Loops','Loops/Degree','MXG')]
    write.csv(t[1:20,], '../tables/2.csv',row.names=F)
}

>>>>

---
jupyter:
  jupytext:
    formats: nb//ipynb,rmd//Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.4.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

@@

```{r}
TabulateSequence <- function(events, possession_sequence) {
    #' Return table starting from row before possession sequence begins, 
    #' and continuing until it ends.

    start <- min(events[events$possession == possession_sequence,]$X)
    stop <- max(events[events$possession == possession_sequence,]$X) + 1
    table <- events[start:stop,]
    df <- data.frame(table$matchPeriod, table$time, table$team, 
                     table$source, table$subEventName, table$possession)
    names(df) <- c('Half', 'Time', 'Team', 'Player', 
                   'Event Description', 'Sequence ID' )

@@

    d <- read.csv('../tables/MXG.csv')

    top <- d[d$Degree>1000,]
    rownames(top) <- c()
    top['Loops/Degree'] <- top$Loops/top$Degree
    top['Goals/Degree'] <- top$Goals/(top$Degree)
    t <- top[c('Player','Team','Degree','Goals',
               'Goals/Degree','Loops','Loops/Degree','MXG')]

    t[,'Goals/Degree'] <- formatC(signif(t[,'Goals/Degree']*100,digits=3), 
                              digits=3,format="fg", flag="#")

    t[,'Goals (% Degree)'] <- paste0(
                            as.character(t$Goals),
                            ' (',
                            as.character(t[,'Goals/Degree']),'%)')

    t$Degree <- as.character(prettyNum(t$Degree, big.mark=',',scientific=F))

    t[,'Loops/Degree'] <- formatC(signif(t[,'Loops/Degree']*100,digits=3), 
                              digits=3,format="fg", flag="#")

    t[,'Loops (% Degree)'] <- paste0(
                            as.character(t$Loops),
                            ' (',
                            as.character(t[,'Loops/Degree']),'%)')

    t <- t[1:20,c(1,2,8,3,10,9)]

    write.csv(t[1:20,], '../tables/Top20.csv',row.names=F)
}

### 2020-07-25 rmd/draft.Rmd
done

Association football is the world's most popular sport [@IPalacios-Huerta2004] and a multibillion dollar industry [@Deloitte2019]. 
Statistical analysis of the sport was pioneered more than fifty years ago [@CReepBenjamin1968], but has only recently begun to make a more widespread impact [@PCintiaEtAl2015]. Such analysis is of potential interest to coaches and scouts working professionally within the game, journalists and media commentators explaining the game, and gamblers and bookmakers putting money on the game. In this paper I explain how Social Network Analysis [@SWassermanEtAl1994] can provide the tools to model a football team's ball-possession as a dynamic probabilistic system, and hence how we can quantify the contribution of each player to the team's goal-scoring capability.

@@

The dataset is provided by WyScout, one of a growing number of companies in the football analytics industry, and is available from FigShare.com [@LPappalardoMassucco2019]. It includes more than three million *event-logs* for every 2017-18 match from the five major European leagues, in JSON format (see Figure 1), with  the details of players, teams, and informative tags (such as the crucial 'Goal') numerically encoded. Each event is associated with an event type (such as 'Shot', 'Pass' or 'Free Kick'), identified with a player, and includes a nested array of the locational *positions* of the event itself and its target.

@@

I used Python to convert this into a *tidy* dataframe [@HWickham2014], which I then analyzed and visualized using R. For the sake of reproducible research [@RPeng2011], all code is attached as appendices, as well as available on Github.

@@

### Exploratory Visualization

We can then visualize the location of the event-log data points in the context of a diagrammatic football pitch, to allow for an immediate and accessible view of the events of any match (see Figures 2 & 3), or particular possession sequence (see Figure 4).

@@

@@

<table><tr><td>
<img src="../figures/EventMapHome.png">
</td><td>
<img src="../figures/EventMapAway.png">
</td></tr></table>

@@

<table><tr><td>
<img src="../figures/PassMapHome.png">
</td><td>
<img src="../figures/PassMapAway.png">
</td></tr></table>

@@

@@

<img src="../figures/PossessionSequence.png">
<figcaption>Figure 4: Pitchmap showing Possession Sequence</figcaption>

@@

We can compare this to a tabulated account of all events from just before the beginning of the possession sequence until its end, and we see that our algorithm has successfully handled the case where an opposition player's unsuccessful challenge interrupts the sequence of events associated with the team in possession.

<figcaption>
Table 1: Events of Possession Sequence
</figcaption>

@@

table1 = pd.read_csv('../tables/1.csv',index_col=0)
print(table1.to_markdown(showindex=False))

@@

<img src="../figures/YouTubeClip.gif" width=800>
<figcaption>Figure 5: YouTube Clip showing successful conclusion of possession sequence
</figcaption>

@@

Social Network Analysis applies the concepts of mathematical *graph theory*.

@@

On an undirected graph, we call the number of edges connecting a node its *degree*; in a directed graph we distinguish between the *indegree* and *outdegree*.

@@

We consider the possession sequences of a football match as *walks* of the football across a social graph in which the nodes are players (considered either individually or collectively), and the connecting edges are defined by the movement of the ball, as it travels from player to player. We can then construct the underlying graph for any particular possession sequence (as in Figure i), or indeed for the whole game.

@@

<img src="../figures/PossessionGraph.png">
<figcaption>Figure 6: Graph of Possession Sequence leading to Goal
</figcaption>

@@

There are several different 'weights' we could assign an edge: the Euclidean distance moved by the ball; the polar angle; the time interval. However, for this particular analysis we focus simply on the network connections between players.

@@

<img src="../figures/AdjacencyMatrixAway.png">

@@

It is increasingly common [@TKnutson2018] to visualize a possession network by positioning each node according to the mean location of the player's touches, and varying the appearance (width, colour or transparency) of the edges to show the frequency of that pass. This swiftly communicates a team's formation, and gives a qualitative sense of a team's strategy.

@@

<table><tr><td>
<img src="../figures/PassingNetworkHome.png">
</td><td>
<img src="../figures/PassingNetworkAway.png">
</td></tr></table>

@@

To consider the team's ball-possession in its dynamic development, rather than merely as a static structure, we also need some definitions from *probability theory*.

@@

A stochastic process has the *Markov property* if $\forall t \ge 0, \mathbb{P}(X_{t+1}=s| X_{0},...,X_{t}) = \mathbb{P}(X_{t+1}=s| X_{t})$; this means that future states are dependent only on the present state, regardless of the previous history of past states. Such a process can be called a Markov process, and we call the probability $\mathbb{P}(X_{t+1}=s_{j}|X_{t}=s_{i})$ the *transition probability* $p_{ij}$ .

@@

<img src="../figures/TransitionMatrixAway.png">

@@

### Calculating Markovian Expected Goals

In Figure 8 we show the heatmap of the Transition Matrix derived from the Adjacency Matrix shown in Figure 6. There we showed only the adjacency counts for the starting eleven, but now we include the subsitutes, as well as treating the 'Opposition' collectively as a single node. Most significantly, we now include a 'Goal' as a node. In particular, in the terms of Markov theory, we treat it as an *absorbing state* which loops back on itself with certain probability. This allows us to calculate the expected *hitting time* of a goal, given that any particular player has the ball.

@@

Thus we can quantify, for each player on the team, the expected number of events there will be before the team scores a goal. We invert this to give a measure of the expected number of goals; and scale it by the team's average number of events per game, we call this measure *Markovian Expected Goals* (MXG). We then scale this by the team's mean expected goals values to give a *Relative MXG* value, so that values can be compared fairly across teams to see which individual players contribute the most to their teams' goalscoring chances.

@@

The

<figcaption>
Table 2: Top Twenty European Players by MXG Score
</figcaption>

@@

MXGtable = pd.read_csv('../tables/2.csv',index_col=0)
print(MXGtable.to_markdown(floatfmt="#.3g"))

>>>>

Association football is the world's most popular sport [@IPalacios-Huerta2004] and a multibillion dollar industry [@Deloitte2019]. 
Statistical analysis of the sport was pioneered more than fifty years ago [@CReepBenjamin1968], but has only recently begun to make a more widespread impact [@PCintiaEtAl2015]. Such analysis is of potential interest to coaches and scouts working professionally within the game, journalists and media commentators explaining the game, and gamblers and bookmakers putting money on the game. In this paper I build upon previous work applying Social Network Analysis [@TGrund2012]
and Markov Modelling [@JPena2014] to football data, and introduce an original metric for quantifying the contribution of each player to a team's goal-scoring capability.

@@

The dataset is provided by WyScout, and is available from FigShare.com [@LPappalardoMassucco2019]. It includes more than three million *event-logs* for every 2017-18 match from the five major European leagues, in JSON format (see Figure 1), with  the details of players, teams, and informative tags (such as the crucial 'Goal') numerically encoded. Each event is associated with an event type (such as 'Shot', 'Pass' or 'Free Kick'), identified with a player, and includes a nested array of the locational *positions* of the event itself and its target.

@@

I used Python to convert this into a *tidy* dataframe [@HWickham2014], which I then analyzed and visualized using R. For the sake of reproducible research [@RPeng2011], all code is attached as appendices.

@@

### Initial Exploration

We begin our exploration of the data by tabulating the frequency of event types (Table 1) and pass types (Table 2), confirming that the *simple pass* is football's primary event.

@@

We can then visualize the location of the event-log data points in the context of a diagrammatic football pitch, to allow for an immediate and accessible view of the events of any match (see Figures 2 & 3).

@@

<table><tr><td>
<img src="figures/EventMapHome.png">
</td><td>
<img src="figures/EventMapAway.png">
</td></tr></table>

@@

<table><tr><td>
<img src="figures/PassMapHome.png">
</td><td>
<img src="figures/PassMapAway.png">
</td></tr></table>

@@

As well as the general overview of a full match's events, we can also consider a particular possession sequence (see Figure 4).

@@

<img src="figures/PossessionSequence.png">
<figcaption>Figure 4: Pitchmap showing Possession Sequence</figcaption>

@@

We can compare this to a tabulated account of all events from just before the beginning of the possession sequence until its end, and we see that our algorithm has successfully handled the case where an opposition player's challenge interrupts the sequence of events associated with the team in possession.

<details open>
<summary>Table 3</summary>
<figure>
----

<figcaption>
Table 3: Events of Possession Sequence
</figcaption>

@@

table = pd.read_csv('../tables/PossessionSequence.csv',index_col=False)
print(table.to_markdown(showindex=False))

@@

<img src="figures/YouTubeClip.gif" width=800>
<figcaption>Figure 5: YouTube Clip showing successful conclusion of possession sequence
</figcaption>

@@

Social Network Analysis [@SWassermanFaust1994] applies the concepts of mathematical *graph theory*.

@@

On an undirected graph, we call the number of edges connecting a node its *degree*. In a directed graph we distinguish between the *indegree* and *outdegree*.

@@

We consider the possession sequences of a football match as *walks* of the football across a social graph in which the nodes are players (considered either individually or collectively), and the connecting edges are defined by the movement of the ball, as it travels from player to player. We can then construct the underlying graph for any particular possession sequence (as in Figure 6), or indeed for the whole game.

@@

<img src="figures/PossessionGraph.png">
<figcaption>Figure 6: Graph of Possession Sequence leading to Goal
</figcaption>

@@

There are several different 'weights' we could assign an edge: the Euclidean distance moved by the ball; the polar angle; the time duration. However, for this particular analysis we focus simply on the network connections between players.

@@

<img src="figures/AdjacencyMatrixAway.png">

@@

It is increasingly common [@TKnutson2018] to visualize a possession network by positioning each node according to the mean location of the player's touches, and varying the appearance (width, colour or transparency) of the edges to show the frequency of that pass (Figure 8). This swiftly communicates a team's formation, and gives a qualitative sense of a team's strategy.

@@

<table><tr><td>
<img src="figures/PassingNetworkHome.png">
</td><td>
<img src="figures/PassingNetworkAway.png">
</td></tr></table>

@@

To consider the team's ball-possession in its dynamic development, rather than merely as a static structure, we also need some definitions from *probability theory* [@RSerfozo2009a].

@@

A stochastic process has the *Markov property* if $\forall t \ge 0, \mathbb{P}(X_{t+1}=s| X_{0},...,X_{t}) = \mathbb{P}(X_{t+1}=s| X_{t})$; this means that future states are dependent only on the present state, regardless of the previous history of past states. Such a process can be called a Markov process [@JNorris1998], and we call the probability $\mathbb{P}(X_{t+1}=s_{j}|X_{t}=s_{i})$ the *transition probability* $p_{ij}$ .

@@

<img src="figures/TransitionMatrixAway.png">

@@

### Calculating 'Markov Expected Goals'

In Figure 9 we show the heatmap of the Transition Matrix derived from the Adjacency Matrix shown in Figure 7. There we showed only the adjacency counts for the starting eleven, but now we include the subsitutes, as well as treating the 'Opposition' collectively as a single node. Most significantly, we now include a 'Goal' as a node. In particular, in the terms of Markov theory, we treat it as an *absorbing state* which loops back on itself with certain probability. This allows us to calculate the expected *hitting time* of a goal, given that any particular player has the ball.

@@

Thus we can quantify, for each player on the team, the expected number of events there will be before the team scores a goal. We invert this to give a measure of the expected number of goals; and standardize it by the team's average number of events per game. We then scale this by the team's mean values, so that we can make a fair comparison across teams to see which individual players contribute the most to their teams' goalscoring chances. We call this measure *Markov Expected Goals* (MXG).

@@

Applying our method to the cumulative season's transition matrix for every team in the dataset, we find that Inter's Icardi is the player who contributed most to the chance of his team scoring (considering only players with degree of at least 1000). His MXG score is almost three times higher than that of Messi, who only just makes it into the Top Twenty (Table 4).

It is instructive to consider Messi's score in order to understand what a high MXG score implies. Of the twenty, Messi has the highest number of goals, the highest number of loops, and the highest (out-)degree, but the very strength of his degree score means that his percentage of goals per degree (which correlates most strongly with MXG) is the lowest.

<details open>
<summary>Table 4</summary>
<figure>
----

<figcaption>
Table 4: Top Twenty Players in European Leagues by MXG Score (2017-18)
</figcaption>

@@

MXGtable = pd.read_csv('../tables/Top20.csv',index_col=0)
print(MXGtable.to_markdown(floatfmt="#.3g"))
